Reproducibility of research during COVID-19: examining the case of population density and the basic reproductive rate from the perspective of spatial analysis

### 2021-09-16 R0-Density-Reanalysis/R0-Density-Reanalysis.tex
Revisions after decision by Geographical Analysis

% Pandoc citation processing
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for calculating minipage widths
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

  \begin{abstract}
  The emergence of the novel SARS-CoV-2 coronavirus and the global
  COVID-19 pandemic has led to explosive growth in scientific research.
  Alas, much of the research in the literature lacks conditions to be
  reproducible, and recent publications on the association between
  population density and the basic reproductive number of SARS-CoV-2 are
  no exception. Relatively few papers share code and data sufficiently,
  which hinders not only verification but additional experimentation. In
  this paper, an example of reproducible research shows the potential of
  spatial analysis for epidemiology research during COVID-19.
  Transparency and openness means that independent researchers can, with
  relatively modest efforts, verify findings and use different
  approaches as appropriate. Given the high stakes of the situation, it
  is essential that scientific findings, on which good policy depends,
  are as robust as possible; as the empirical example shows,
  reproducibility is one of the keys to ensure this.
  \end{abstract}

@@

Given the ruinous human and economic cost of the pandemic, there has
been a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research has termed the
deluge of COVID-19 publications a ``carnage of substandard research''
(Bramstedt, 2020). Part of the challenge of maintaining quality
standards in published research is that, despite an abundance of
recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
practice reproducibility has remained a lofty and somewhat aspirational
goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
et al., 2018), and the situation does not appear to have changed
substantially since (Gustot, 2020; Sumner et al., 2020).

The push for open data and software, along with more strenuous efforts
towards open, reproducible research, is simply a continuation of
long-standing scientific practices of independent verification. Despite
the (at times disproportionate) attention that high profile scandals in
science tend to elicit in the media, science as a collective endeavor is
remarkable for being a self-correcting enterprise, one with built-in
mechanisms and incentives to weed out erroneous ideas. Over the long
term, facts tend to prevail in science. At stake is the shorter-term
impacts that research may have in other spheres of economic and social
life. The case of economists Reinhart and Rogoff comes to mind: by the
time the inaccuracies and errors in their research were uncovered (see
Herndon et al., 2014), their claims about debt and economic growth had
already been seized by policy-makers on both sides of the Atlantic to
justify austerity policies in the aftermath of the Great Recession of
2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
  ``Reinhart--Rogoff may have had more immediate influence on public
  debate than any previous paper in the history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years (Basu et al., 2017).

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic (Florida et al., 2020); the implications of the pandemic for
urban planning, design, and management are the topic of ongoing research
(Sharifi and Khavarian-Garmsir, 2020). The fact that the worst of the
pandemic was initially felt in dense population centers such as Wuhan,
Milan, Madrid, and New York, unleashed a torrent of research into the
associations between density and the spread of the pandemic. he answers
to some important questions hang on the results of these research
efforts. For example, are lower density regions safer from the pandemic?
Are de-densification policies warranted, even if just in the short term?
And in the longer term, will the risks of life in high density regions
presage a flight from cities? What are the implications of the pandemic
for future urban planning and practice? Over the past year, numerous
papers have sought to throw light on the underlying issue of density and
the pandemic; nonetheless the results, as will be detailed next, remain
mixed. Further, to complicate matters, precious few of these studies
appear to be sufficiently open to support independent verification.

The objective of this paper is to illustrate the importance of
reproducibility in research in the context of the flood of COVID-19
papers. To this end, a recent study by Sy et al. (2021) is chosen as an
example of reproducible research. The objective is not to malign the
analysis of these researchers, but rather to demonstrate the value of
openness to allow for independent verification and further analysis.
Open data and open code mean that an independent researcher can, with
only modest efforts, not only verify the findings reported, but also
examine the same data from a perspective which may not have been
available to the original researchers due to differences in disciplinary
perspectives, methodological traditions, and/or training, among other
possible factors. The example, which shows consequential changes in the
conclusions reached by different analyses, should serve as a call to
researchers to redouble their efforts to increase transparency and
reproducibility in research. The present paper, in addition, aims to
show how data can be packaged in well-documented, shareable units, and
code can be embedded into self-contained documents suitable for review
and independent verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, will be available in a public
repository\footnote{For peer-review purposes, the data package and code
  are currently in an anonymous Drive folder:
  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.

@@

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density may have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density (see for example the research on urban fields and
time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
Brown, 1970). Mathematically, models of exposure and contagion indicate
that higher densities can catalyze the transmission of contagious
diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
intuitive and likely at the root of messages, by some figures in
positions of authority, that regions with sparse population densities
faced lower risks from the pandemic\footnote{Governor Kristi Noem of
  South Dakota, for example, claimed that sparse population density
  allowed her state to face the pandemic down without the need for
  strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
mathematical models of contagion are valid at small-to-medium spatial
scales (and presumably, small temporal scales too, such as time spent in
restaurants, concert halls, cruises), and the results do not necessarily
transfer to larger spatial units and different time scales. There are
solid reasons for this: while in a restaurant, one can hardly avoid
being in proximity to other customers. On the other hand, a person can
choose to (or be forced to as a matter of policy) not go to a restaurant
in the first place. Nonetheless, the idea that high density correlates
with high transmission is so seemingly sensible that it is often taken
for granted even at larger scales (e.g., Cruz et al., 2020; Micallef et
al., 2020). At larger scales, however, there exists the possibility of
behavioral adaptations, which are difficult to capture in the
mechanistic framework of differential equations (or can be missing in
agent-based models, e.g., Gomez et al., 2021); these adaptations, in
fact, can be a key aspect of disease transmission.

@@

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. (2020) who did not find evidence of
significant correlation between population density and confirmed number
of cases per day \emph{in conditions of lockdown} in China. This finding
echoes the results of Paez et al. (2020), who in their study of
provinces in Spain reported non-significant associations between
population density and infection rates in the early days of the first
wave of COVID-19, and negative significant associations in the later
part of the first lockdown. Similarly, (2020) found zero or negative
associations between population density and infection numbers/deaths by
country. Fielding-Miller et al. (2020) contrast their finding about
rural counties with a negative relationship between COVID-19 deaths and
population density in urban counties in the US. For their part, in their
investigation of doubling time, White and Hébert-Dufresne (2020)
identified a negative and significant correlation between population
density and doubling time in US states. Likewise, (2021) found a small
negative (and significant) association between population density and
COVID-19 morbidity in districts in Tehran. Finally, two of the most
complete studies in the US (by Hamidi et al. 2020a, and 2020b) used an
extensive set of controls to find negative and significant correlations
between density and COVID-19 cases and fatalities at the level of
counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al.~{[}(2021); hereafter SWN{]} would immediately grab the attention
of a researcher with expertise in spatial analysis.

@@

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. For their multivariate
analysis, SWN used mixed linear models. This is a reasonable modelling
choice: \(R_0\) is an interval-ratio variable that is suitably modeled
using linear regression; further, as SWN note there is a likelihood that
the process in not independent ``among counties within each state,
potentially due to variable resource allocation and differing health
systems across states'' (p.~3). A mixed linear model accounts for this
by introducing random components; in the case of SWN, these are random
intercepts at the state level. SWN estimated various models with
different combinations of variables, including median household income
and prevalence of travel by private transportation. These are sensible
controls, given potential variations in behavior: people in more
affluent counties may have greater opportunities to work from home, and
use of private transportation reduces contact with strangers. Moreover,
they also conducted various sensitivity analyses. After these efforts,
SWN concluded that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

One salient aspect of the analysis in SWN is that the basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made modelling decisions, in this case basing their
analysis only on counties with valid observations. A modeler with
expertise in spatial analysis would likely ask some of the following
questions on reading SWN's paper: how were missing counties treated?
What are the implications of the spatial sampling framework used in the
analysis? Is it possible to spatially interpolate the missing
observations? Was there spatial residual autocorrelation in the models,
or was the use of mixed models sufficient to capture spatial
dependencies? These questions are relevant and their implications
important. Fortunately, SWN are an example of a reasonably open,
reproducible research product: their paper is accompanied by (most of)
the data and (most of) the code used in the analysis. This means that an
independent expert can, with only a moderate investment of time and
effort, reproduce the results in the paper, as well as ask additional
questions.

@@

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
2020). But in many other cases the data are publicly available, which in
fact has commonly been the case with population-level COVID-19
information. Typically the provenance of the data is documented, but in
numerous studies the data themselves are not shared (Amadu et al., 2021;
Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
whether a graduate student or a seasoned scientist, collecting,
organizing, and preparing data for a project can take a substantial
amount of time. Pointing to the sources of data, even when these sources
are public, is a small step towards reproducibility-but only a very
small one. Faced with the prospect of having to recreate a data set from
raw sources is probably sufficient to dissuade all but the most
dedicated (or stubborn) researcher from independent verification. This
is true even if part of the data are shared (e.g., Wong and Li, 2020).
In other cases, data are shared, but the processes followed in the
preparation of the data are not fully documented (Ahmad et al., 2020;
Skórka et al., 2020). These processes matter, as shown by the errors in
the spreadsheets of Reinhart and Rogoff (Herndon et al., 2014) and the
data of biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions\footnote{\url{https://doi.org/10.1038/d41586-020-00287-y}}.
Another situation is when papers share well-documented data, but fail to
provide the code used in the analysis (Noury et al., 2021; Pequeno et
al., 2020; Wang et al., 2021). Making code available only ``on demand''
(e.g., Brandtner et al., 2021) is an unnecessary barrier when most
journals offer the facility to share supplemental materials online. Then
there are those papers that more closely comply with reproducibility
standards, and share well-documented processes and data, as well as the
code used in any analyses reported (Feyman et al., 2020; Paez et al.,
2020; Stephens et al., 2021; Sy et al., 2021; White and Hébert-Dufresne,
2020). Even in this case, the pressure to publish ``new findings''
instead of replication studies can act as a deterrent, perhaps
particularly for younger researchers\footnote{The present paper was desk
  rejected by three journals that had previously published research on
  population density and the spread of COVID-19; in one case, the paper
  was too opinionated for the journal, in the other two cases, the paper
  was not a ``good fit'' despite dealing with a nearly identical issue
  as papers previously published in said journals. This does not inspire
  much confidence in the commitment of journals to reproducibility in
  research.}.

In the following sections, the analysis of RWN is reproduced, some
relevant questions from the perspective of an independent researcher
with expertise in spatial analysis are asked, and the data are
reanalyzed.

@@

SWN examined the association between the basic reproductive number of
COVID-19 and population density. The basic reproductive number \(R_0\)
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, \(R_0\)
measures how many new infections each infections begets. Infectious
disease outbreaks generally tend to die out when \(R_0<1\), and to grow
when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
number of cases to be able to assume that there is community
transmission of the pathogen. Accordingly, SWN based their analysis only
on counties that had at least 25 cases or more at the end of the
exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
Rico, and island territories.

@@

Table \ref{tab:swn-results} reproduces the first three models of SWN
(the fourth model did not have any significant variables; see Table 1 in
SWN). It is possible to verify that the results match, with only the
minor (and irrelevant) exception of the magnitude of the coefficient for
travel by private transportation, which is due to a difference in the
input (here the variable is changed to one percent units, instead of the
ten percent units used by SWN). The mixed linear model gives random
intercepts (i.e., the intercept is a random variable), and the standard
deviation is reported in the fourth row of Table \ref{tab:swn-results}.
It is useful to map the random intercepts: as seen in Figure
\ref{fig:random-terms-map}, other things being equal, counties in Texas
tend to have somewhat lower values of \(R_0\) (i.e., a negative random
intercept), whereas counties in South Dakota tend to have higher values
of \(R_0\). The key of the analysis, after extensive sensitivity
analysis, is a robust finding that population density has a positive
association with the basic reproductive number. But does it?

@@

@@

How relevant are the differences between the various model
specifications presented above? Figure \ref{fig:comparison-results}
shows the relationship between density and \(R_0\) implied by SWN Model
3, Selection Model 2, and Selection Model 3. The left panel of the
figure shows the non-linear but monotonic relationship implied by SWN
Model 1. The conclusion is that at higher densities, \(R_0\) is
\emph{always} higher. The two panels on the right, in contrast, shows
that Selection Model 2 and Selection Model 3 coincide that \(R_0\) tends
to increase as density grows. This continues until a density of
approximately 2.9 (1,000 people per sq.km). At higher densities than
that the relationship between density and \(R_0\) begins to weaken, and
the relationship becomes negative at densities higher than approximately
5.7 (1,000 people per sq.km).

>>>>

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

  \begin{abstract}
  The emergence of the novel SARS-CoV-2 coronavirus and the global
  COVID-19 pandemic in 2019 led to explosive growth in scientific
  research. Alas, much of the research in the literature lacks
  conditions to be reproducible, and recent publications on the
  association between population density and the basic reproductive
  number of SARS-CoV-2 are no exception. Relatively few papers share
  code and data sufficiently, which hinders not only verification but
  additional experimentation. In this paper, an example of reproducible
  research shows the potential of spatial analysis for epidemiology
  research during COVID-19. Transparency and openness means that
  independent researchers can, with only modest efforts, verify findings
  and use different approaches as appropriate. Given the high stakes of
  the situation, it is essential that scientific findings, on which good
  policy depends, are as robust as possible; as the empirical example
  shows, reproducibility is one of the keys to ensure this.
  \end{abstract}

@@

Given the ruinous human and economic cost of the pandemic, there has
been a natural tension in the scientific community between the need to
publish research results quickly and the imperative to maintain
consistently high quality standards in scientific reporting; indeed, a
call for maintaining the standards in published research termed the
deluge of COVID-19 publications a ``carnage of substandard research''
(Bramstedt, 2020). Part of the challenge of maintaining quality
standards in published research is that, despite an abundance of
recommendations and guidelines (e.g., Broggini et al., 2017; Brunsdon
and Comber, 2020; Ince et al., 2012; Ioannidis et al., 2014), in
practice reproducibility has remained a lofty and somewhat aspirational
goal (Konkol et al., 2019; Konkol and Kray, 2019). As reported in the
literature, only a woefully small proportion of published research was
actually reproducible before the pandemic (Iqbal et al., 2016; Stodden
et al., 2018), and the situation does not appear to have changed
substantially since (Gustot, 2020; Sumner et al., 2020).

The push for open software and data (Arribas-Bel et al., 2021; e.g.,
Bivand, 2020), along with more strenuous efforts towards open,
reproducible research, is simply a continuation of long-standing
scientific practices of independent verification. Despite the (at times
disproportionate) attention that high profile scandals in science tend
to elicit in the media, science as a collective endeavor is remarkable
for being a self-correcting enterprise, one with built-in mechanisms and
incentives to weed out erroneous ideas. Over the long term, facts tend
to prevail in science. At stake is the shorter-term impacts that
research may have in other spheres of economic and social life. The case
of economists Reinhart and Rogoff comes to mind: by the time the
inaccuracies and errors in their research were uncovered (see Herndon et
al., 2014), their claims about debt and economic growth had already been
seized by policy-makers on both sides of the Atlantic to justify
austerity policies in the aftermath of the Great Recession of
2007-2009\footnote{Nobel Prize in Economics Paul Krugman noted that
  ``Reinhart--Rogoff may have had more immediate influence on public
  debate than any previous paper in the history of economics''
  \url{https://www.nybooks.com/articles/2013/06/06/how-case-austerity-has-crumbled/?pagination=false}}.
As later research has demonstrated, those policies cast a long shadow,
and their sequels continued to be felt for years (Basu et al., 2017).

In the context of COVID-19, a topic that has grabbed the imagination of
numerous thinkers has been the prospect of life in cities after the
pandemic (e.g., Florida et al., 2020); as a result, the implications of
the pandemic for urban planning, design, and management are the topic of
ongoing research (e.g., Sharifi and Khavarian-Garmsir, 2020). The fact
that the worst of the pandemic was initially felt in dense population
centers such as Wuhan, Milan, Madrid, and New York, unleashed a torrent
of research into the associations between density and the spread of the
pandemic. The answers to some important questions hang on the results of
these research efforts. For example, are lower density regions safer
from the pandemic? Are de-densification policies warranted, even if just
in the short term? In the longer term, will the risks of life in high
density regions presage a flight from cities? And, what are the
implications of the pandemic for future urban planning and practice?
Over the past year, numerous papers have sought to throw light on the
underlying issue of density and the pandemic; nonetheless the results,
as will be detailed next, remain mixed. Further, to complicate matters,
precious few of these studies appear to be sufficiently open to support
independent verification.

The objective of this paper is to illustrate the importance of
reproducibility in research in the context of the flood of COVID-19
papers. For this, I focus on a recent study by Sy et al. (2021) that
examined the correlation between the basic reproductive number of
COVID-19, \(R_0\), and population density. The basic reproductive number
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, it measures
how many new infections each infections begets. The paper of Sy et al.
(2021) was selected for being, in the literature examined, almost alone
in supporting reproducible research. Accordingly, I wish to be clear
that my objective in singling their work for discussion is not to malign
their efforts, but rather to demonstrate how open and reproducible
research efforts can greatly help to accelerate discovery. More
concretely, open data and open code mean that an independent researcher
can, with only modest efforts, not only verify the findings reported,
but also examine the same data from a perspective which may not have
been available to the original researchers due to differences in
disciplinary perspectives, methodological traditions, and/or training,
among other possible factors. The example, which shows consequential
changes in the conclusions reached by different analyses, should serve
as a call to researchers to redouble their efforts to increase
transparency and reproducibility in their research. In this spirit, the
present paper also aims to show how data can be packaged in
well-documented, shareable units, and code can be embedded into
self-contained documents suitable for review and independent
verification. The source for this paper is an
\href{http://rmarkdown.rstudio.com}{R Markdown} document which, along
with the data package, will be available in a public
repository\footnote{For peer-review purposes, the data package and code
  are currently in an anonymous Drive folder:
  \url{https://drive.google.com/drive/folders/1cT6tcUc1pJ4aT5ajQ0emO0lyS46P8Ige?usp=sharing}}.

@@

The concern with population density and the spread of the virus during
the COVID-19 pandemic was fueled, at least in part, by dramatic scenes
seen in real-time around the world from large urban centers such as
Wuhan, Milan, Madrid, and New York. In theory, there are good reasons to
believe that higher density could have a positive association with the
transmission of a contagious virus. It has long been known that the
potential for inter-personal contact is greater in regions with higher
density (see for example the research on urban fields and
time-geography, including Farber and Páez, 2011; Moore, 1970; Moore and
Brown, 1970). Mathematically, models of exposure and contagion indicate
that higher densities can catalyze the transmission of contagious
diseases (Li et al., 2018; Rocklöv and Sjödin, 2020). The idea is
intuitive and likely at the root of messages, by some figures in
positions of authority, that regions with sparse population densities
faced lower risks from the pandemic\footnote{Governor Kristi Noem of
  South Dakota, for example, claimed that sparse population density
  allowed her state to face the pandemic down without the need for
  strict policy interventions
  \url{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}}.

As Rocklöv and Sjödin (Rocklöv and Sjödin, 2020) note, however,
mathematical models of contagion are valid at small-to-medium spaces
(and presumably, smaller time intervals too, such as time spent in
restaurants, concert halls, cruises), and the results do not necessarily
transfer to larger spatial units and longer time periods. There are
solid reasons for this: while in a restaurant, one can hardly avoid
being in proximity to other customers. On the other hand, a person can
choose to (or be forced to as a matter of policy) not go to a restaurant
in the first place. Nonetheless, the idea that high density correlates
with high transmission is so seemingly sensible that it is often taken
for granted even at the scale of large spaces (e.g., Cruz et al., 2020;
Micallef et al., 2020). In such conditions, however, there exists the
possibility of behavioral adaptations, which are difficult to capture in
the mechanistic framework of differential equations (or can be missing
in agent-based models, e.g., Gomez et al., 2021); these adaptations, in
fact, can be a key aspect of disease transmission.

@@

On the flip side, a number of studies report non-significant or negative
associations between population density and COVID-19 outcomes. This
includes the research of Sun et al. (2020) who did not find evidence of
significant correlation between population density and confirmed number
of cases per day \emph{in conditions of lockdown} in China. This finding
echoes the results of Paez et al. (2020), who in their study of
provinces in Spain reported non-significant associations between
population density and infection rates in the early days of the first
wave of COVID-19, and negative significant associations in the later
part of the first lockdown. Similarly, Skórka et al. (2020) found zero
or negative associations between population density and infection
numbers/deaths by country. Fielding-Miller et al. (2020) contrast their
finding about rural counties with a negative relationship between
COVID-19 deaths and population density in urban counties in the US. For
their part, in their investigation of doubling time, White and
Hébert-Dufresne (2020) identified a negative and significant correlation
between population density and doubling time in US states. Likewise,
Khavarian-Garmsir et al. (2021) found a small negative (and significant)
association between population density and COVID-19 morbidity in
districts in Tehran. Finally, two of the most complete studies in the
US, by Hamidi et al. (2020a) and Hamidi et al. (2020b), used an
extensive set of controls to find negative and significant correlations
between density and COVID-19 cases and fatalities at the level of
counties in the US.

As can be seen, these studies are implemented at different scales in
different regions of the world. They also use a range of techniques,
from correlation analysis, to multivariate regression, spatial
regressions, and machine learning techniques. This is natural and to be
expected: individual researchers have only limited time and expertise.
This is why reproducibility is important. To pick an example (which will
be further elaborated in later sections of this paper), the study of Sy
et al. (2021), hereafter referred to as SWN, would immediately grab the
attention of a researcher with expertise in spatial analysis.

@@

SWN investigated the basic reproductive number of COVID-19 in US
counties, and its association with population density, median household
income, and prevalence of private mobility. For their multivariate
analysis, SWN used mixed linear models. This is an appropriate modelling
choice: \(R_0\) is an interval-ratio variable that is suitably modeled
using linear regression; further, as SWN note there is a likelihood that
the process in not independent ``among counties within each state,
potentially due to variable resource allocation and differing health
systems across states'' (p.~3). A mixed linear model accounts for this
by introducing random components; in the case of SWN, these are random
intercepts at the state level. SWN estimated various models with
different combinations of variables, including median household income
and prevalence of travel by private transportation. These controls help
to account for potential variations in behavior: people in more affluent
counties may have greater opportunities to work from home, and use of
private transportation reduces contact with strangers. Moreover, they
also conducted various sensitivity analyses. After these efforts, SWN
concluded that there is a positive association between the basic
reproductive number and population density at the level of counties in
the US.

One salient aspect of the analysis in SWN is that the basic reproductive
number can only be calculated reliably with a minimum number of cases,
and a large number of counties did not meet such threshold. As
researchers do, SWN made modelling decisions, in this case basing their
analysis only on counties with valid observations. A modeler with
expertise in spatial analysis would likely ask some of the following
questions on reading SWN's paper: how were missing counties treated?
What are the implications of the spatial sampling framework used in the
analysis? Is it possible to spatially interpolate the missing
observations? Was there spatial residual autocorrelation in the models,
or was the use of mixed models sufficient to capture spatial
dependencies? These questions are relevant and their implications
important. Fortunately, SWN are an example of a reasonably open,
reproducible research product: their paper is accompanied by (most of)
the data and (most of) the code used in the analysis. This means that an
independent researcher can, with only a moderate investment of time and
effort, reproduce the results in the paper, as well as ask additional
questions.

@@

There are various reasons why a project can fail to be reproducible. In
some cases, there might be legitimate reasons to withhold the data,
perhaps due to confidentiality and privacy reasons (e.g., Lee et al.,
2020). But in many other cases the data are publicly available, which in
fact has commonly been the case with population-level COVID-19
information. Typically the provenance of the data is documented, but in
numerous studies the data themselves are not shared (Amadu et al., 2021;
Bhadra et al., 2021; Cruz et al., 2020; Feng et al., 2020;
Fielding-Miller et al., 2020; Hamidi et al., 2020a, 2020b; Inbaraj et
al., 2021; Souris and Gonzalez, 2020). As any researcher can attest,
collecting, organizing, and preparing data for a project can take a
substantial amount of time. Pointing to the sources of data, even when
these sources are public, is a small step towards reproducibility-but
only a very small one. Faced with the prospect of having to recreate a
data set from raw sources is probably sufficient to dissuade all but the
most dedicated (or stubborn) researcher from independent verification.
This is true even if part of the data are shared (e.g., Wong and Li,
2020). In other cases, data are shared, but the processes followed in
the preparation of the data are not fully documented (Ahmad et al.,
2020; Skórka et al., 2020). These processes matter, as shown by the
errors in the spreadsheets of Reinhart and Rogoff (see Herndon et al.,
2014 for the discovery of these errors), as well as by the data of
biologist Jonathan Pruitt that led to an ``avalanche'' of paper
retractions (see Viglione, 2020). Another situation is when papers share
well-documented data, but fail to provide the code used in the analysis
(Noury et al., 2021; Pequeno et al., 2020; Wang et al., 2021). Making
code available only ``on demand'' (e.g., Brandtner et al., 2021) is an
unnecessary barrier when most journals offer the facility to share
supplemental materials online. Then there are those papers that more
closely comply with reproducibility standards, and share well-documented
processes and data, as well as the code used in any analyses reported
(Feyman et al., 2020; Paez et al., 2020; Stephens et al., 2021; Sy et
al., 2021; White and Hébert-Dufresne, 2020). Even in this case, the
pressure to publish ``new findings'' instead of replication studies can
act as a deterrent, perhaps particularly for younger
researchers\footnote{The present paper was desk rejected by three
  journals that had previously published research on population density
  and the spread of COVID-19; in one case, the paper was too opinionated
  for the journal, in the other two cases, the paper was not a ``good
  fit'' despite dealing with a nearly identical issue as papers
  previously published in said journals.}.

In the following sections, the analysis of SWN is reproduced, some
relevant questions from the perspective of an independent researcher
with expertise in spatial analysis are asked, and the data are
reanalyzed.

@@

SWN examined the association between the basic reproductive number of
COVID-19 and population density. The basic reproductive number \(R_0\)
is a summary measure of contact rates, probability of transmission of a
pathogen, and duration of infectiousness. In rough terms, \(R_0\)
measures how many new infections each infections begets. Infectious
disease outbreaks generally tend to die out when \(R_0<1\), and to grow
when \(R_0>1\). Reliable calculation of \(R_0\) requires a minimum
number of cases to be able to assume that there is community
transmission of the pathogen. Accordingly, SWN based their analysis only
on counties that had at least 25 cases or more at the end of the
exponential growth phase (see Fig. \ref{fig:R0-map}). Their final sample
included 1,151 counties in the US, including in Alaska, Hawaii, Puerto
Rico, and island territories. SWN used COVID-19 data collected by the
New York Times and made available (with versioning) in a GitHub
repository\footnote{https://github.com/nytimes/covid-19-data}. For each
county, SWN assumed that the exponential growth period began one week
prior to the second daily increase in cases, and assumed that the period
of exponential growth lasted approximately 18 days.

@@

Table \ref{tab:swn-results} reproduces the first three models of SWN
(the fourth model did not have any significant variables; see Table 1 in
SWN). It is possible to verify that the results match, with only the
minor (and irrelevant) exception of the magnitude of the coefficient for
travel by private transportation, which is due to a difference in the
input (here the variable is changed to one percent units, instead of the
ten percent units used by SWN). The mixed linear model gives random
intercepts (i.e., the intercept is a random variable), and the standard
deviation is reported in the fifth row of Table \ref{tab:swn-results}.
It is useful to map the random intercepts: as seen in Figure
\ref{fig:random-terms-map}, other things being equal, counties in Texas
tend to have somewhat lower values of \(R_0\) (i.e., a negative random
intercept), whereas counties in South Dakota tend to have higher values
of \(R_0\). The key of the analysis, after extensive sensitivity
analysis, is a robust finding that population density has a positive
association with the basic reproductive number. But does it?

@@

Clearly, the various models display some intriguing differences; but how
relevant are said differences from a more substantive standpoint? Figure
\ref{fig:comparison-results} shows the relationship between density and
\(R_0\) implied by SWN Model 3, Selection Model 2, and Selection Model
3. The left panel of the figure shows the non-linear but monotonic
relationship implied by SWN Model 1. The conclusion is that at higher
densities, \(R_0\) is \emph{always} higher. The two panels on the right,
in contrast, shows that Selection Model 2 and Selection Model 3 coincide
that \(R_0\) tends to increase as density grows. This continues until a
density of approximately 2.9 (1,000 people per sq.km). At higher
densities than that the relationship between density and \(R_0\) begins
to weaken, and the relationship becomes negative at densities higher
than approximately 5.7 (1,000 people per sq.km).

@@

It is worth at this point to recall Cressie's dictum about modelling:
``{[}w{]}hat is one person's mean structure could be another person's
correlation structure'' (Cressie, 1989, p. 201). There are almost always
multiple ways to approach a modelling situation, as lively illustrated
by a recent paper that reports the results of a crowdsourced modelling
experiment (Schweinsberg et al., 2021). In the present case, we would
argue that spatial sampling is an important aspect of the modeling
process. Importantly, by adopting high reproducibility standards, SWN
made a valuable contribution to the collective enterprise of seeking
knowledge. Their effort, and subsequent efforts to validate and expand
on their work, can potentially contribute to provide clarity to ongoing
conversations about the relevance of density and the spread of COVID-19.

In particular, it is noteworthy that a sample selection model with a
different variable transformation does not lend support to the thesis
that higher density is \emph{always} associated with a greater risk of
spread of the virus {[}in Wong and Li's words, ``\,`Density is destiny'
is probably an overstatement''; (2020){]}. At the same time, the results
presented here also stand in contrast to the findings of Hamidi et al.,
who found that higher density was either not significantly associated
with the rate of the virus in a cross-sectional study (Hamidi et al.,
2020b), or was negatively associated with it in a longitudinal setting
{[}Hamidi et al. (2020a). In this sense, the conclusion that density
does not aggravate the pandemic may have been somewhat premature;
instead, reanalysis of the data of SWN suggests that Fielding-Miller et
al. (2020) might be onto something with respect to the difference
between rural and urban counties. More generally, there is no doubt that
in population-level studies density is indicative of proximity, but it
also potentially is a proxy for adaptive behavior. And it is possible
that the determining factor during COVID-19, at least in the US, has
been variations in perceptions of the risks associated with contagion
(Chauhan et al., 2021), and subsequent compensations in behavior in more
and less dense regions.
